% =========================================================================
% GRAFT — working draft. DO NOT SUBMIT until:
%   (1) every [TO FILL] table is completed with real experiments,
%   (2) every citation is re-verified against arXiv/DBLP,
%   (3) a fresh literature sweep (last 6 months) confirms novelty claims,
%   (4) the text is revised in the author's own voice.
% See INTEGRITY_CHECKLIST.md in this folder.
% Retarget to the venue template (NeurIPS/CVPR/WACV .sty) before submission.
% =========================================================================
\documentclass[10pt,twocolumn]{article}
\usepackage[margin=0.75in,columnsep=0.25in]{geometry}
\usepackage{mathptmx}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}

\newcommand{\tofill}[1]{\textcolor{red}{\textbf{[TO FILL: #1]}}}
\newcommand{\real}{\mathrm{real}}
\newcommand{\fake}{\mathrm{ai}}

\title{\textbf{GRAFT: Gradient-Field Representations and Frozen-Backbone\\
Fusion for Robust Detection of AI-Generated Images}}
\author{Jitendra Jha\\
\small Independent Researcher\\
\small \texttt{github.com/JitendraJha98/fakeradar}}
\date{Working draft --- \today}

\begin{document}
\maketitle

\begin{center}
\fbox{\parbox{0.95\columnwidth}{\small\textcolor{red}{\textbf{WORKING DRAFT.}}
Method, protocol, and related work are complete; all experimental numbers are
pending and marked \tofill{}. This document must not be submitted or
distributed as a completed study until every placeholder is replaced with
results reproduced by the released code.}}
\end{center}

\begin{abstract}
Detectors of AI-generated images fail along two axes that matter in
deployment: they collapse on images from \emph{unseen} generators, and they
collapse when images traverse real transport chains (JPEG re-encoding,
resizing, platform re-processing). We present GRAFT, a detection architecture
that fuses two families of evidence with complementary failure modes: a
\emph{learnable gradient-field branch} operating on native-resolution crops
--- combining steganalysis-style high-pass residuals, an up-sampling
(NPR-style) residual, and, novelly in this context, the \emph{structure-tensor
coherence} of the image gradient field --- with \emph{frozen}
vision--language-model features, whose resistance to generator overfitting is
preserved precisely because the backbone is never fine-tuned. A gated fusion
head with per-branch deep supervision yields a single temperature-calibrated
probability together with interpretable per-branch scores, and an explicit
abstention band. Training passes every image through a simulated transport
chain, making robustness a training-time property rather than an evaluation
afterthought. Under the strictest GenImage protocol (train on Stable Diffusion
v1.4 only, test on eight generators), GRAFT attains \tofill{mean AUROC},
retaining \tofill{AUROC at JPEG quality 50} under recompression, versus
\tofill{best baseline}. Code, weights, and the full evaluation harness are
released.
\end{abstract}

\section{Introduction}
Photorealistic image generators --- latent diffusion systems, autoregressive
image models, and their commercial descendants --- have made fabricated
imagery cheap at consumer scale. The forensic response, a decade of
increasingly capable ``real vs.\ generated'' classifiers, suffers from a
well-documented pair of gaps between benchmark and deployment. First,
\emph{generator generalization}: a detector trained on one generator family
tends to key on that family's idiosyncratic artifacts and degrades sharply on
newer models~\cite{wang2020cnn,gragnaniello2021gan,zhu2023genimage}. Second,
\emph{transport robustness}: the high-frequency traces most detectors exploit
are attenuated or destroyed by the compression--resize pipelines of messaging
and social platforms~\cite{gragnaniello2021gan,corvi2023detection}, and
adversarial post-processing can defeat detectors with theoretical
backing~\cite{saberi2024robustness}.

The literature contains two strong but largely separate responses. One line
makes \emph{low-level traces} more generator-agnostic: gradients of the image
with respect to a pretrained model (LGrad~\cite{tan2023lgrad}), local
up-sampling residuals (NPR~\cite{tan2024npr}), and spectral
statistics~\cite{durall2020watch,frank2020leveraging}. A second line abandons
low-level traces altogether and reads \emph{frozen foundation-model feature
spaces}, in which generated images cluster even for unseen generators
(UniversalFakeDetect~\cite{ojha2023universal}; CLIP-based
successors~\cite{cozzolino2024raising,liu2024fatformer}). These families fail
differently: low-level branches are strongest on clean, native-resolution
inputs and degrade under recompression; frozen-feature branches survive
transport better but are weaker at separating high-fidelity images whose
semantics are unremarkable. Hybrid designs are beginning to appear
\cite{yan2025sanity}, but --- to our knowledge, as of this draft --- no
published method fuses a \emph{learnable gradient-field representation} with
a \emph{frozen} vision--language backbone under a \emph{robustness-first}
training protocol with calibrated, abstaining outputs. GRAFT is that design,
released as a maintained open-source system rather than a one-off research
artifact.

Our contributions:
\begin{enumerate}\itemsep2pt
  \item \textbf{A gradient-field forensic representation} stacking SRM
  high-pass residuals~\cite{fridrich2012rich}, an NPR-style up-sampling
  residual~\cite{tan2024npr}, gradient magnitude, and --- new in this context
  --- the structure-tensor \emph{coherence} of the gradient field, an
  explicit, resolution-preserving measure of the anomalous local gradient
  alignment of synthetic texture (\S\ref{sec:gradient}).
  \item \textbf{Gated frozen-backbone fusion} with per-branch deep
  supervision: complementary evidence is combined by a learned gate whose
  weights, together with per-branch probabilities, are exposed as an
  interpretability signal; the semantic backbone stays frozen to preserve its
  cross-generator behavior (\S\ref{sec:fusion}).
  \item \textbf{A robustness-first protocol}: transport-chain augmentation at
  training time; native-resolution multi-crop inference that never resamples
  the evidence; temperature-calibrated probabilities with an abstention band;
  and mandatory reporting of perturbed-set metrics, TPR at fixed low FPR, and
  calibration error alongside clean AUROC (\S\ref{sec:protocol}).
  \item \textbf{Reproducible release}: training, evaluation, and robustness
  harnesses, plus weights, as the open-source system \texttt{fakeradar}.
\end{enumerate}

\section{Related Work}
\paragraph{CNN classifiers and their limits.}
Wang et al.~\cite{wang2020cnn} showed a ResNet trained on one GAN transfers
surprisingly well across CNN generators given aggressive augmentation --- and
established the now-standard caveat that this transfer is fragile.
Gragnaniello et al.~\cite{gragnaniello2021gan} documented how quickly such
detectors decay under resizing and compression; Corvi et
al.~\cite{corvi2023detection} extended the critical analysis to diffusion
models.

\paragraph{Gradient- and residual-domain detection.}
Steganalysis rich models~\cite{fridrich2012rich} established fixed high-pass
residuals as content-suppressing noise descriptors. LGrad~\cite{tan2023lgrad}
converts images into gradients of a pretrained CNN, learning on a
generalized artifact representation; NPR~\cite{tan2024npr} shows that a
simple neighboring-pixel relation targeting up-sampling operators
generalizes broadly across CNN-based generators. GRAFT's gradient branch is
a superset of these signals and adds the structure-tensor coherence channel;
unlike LGrad/NPR it is trained \emph{jointly} with, and gated against,
frozen semantic features.

\paragraph{Frequency-domain detection.}
Up-convolutions imprint spectral anomalies~\cite{durall2020watch}, and
frequency inputs alone support strong detection~\cite{frank2020leveraging}.
We retain an azimuthally averaged log-power spectrum as a compact,
cheap branch rather than a stand-alone detector.

\paragraph{Frozen foundation-model features.}
Ojha et al.~\cite{ojha2023universal} demonstrated that nearest-neighbor and
linear probes in frozen CLIP~\cite{radford2021clip} space generalize across
generator families where end-to-end CNNs fail; Cozzolino et
al.~\cite{cozzolino2024raising} and FatFormer~\cite{liu2024fatformer}
strengthen this line. We deliberately keep this branch frozen and lightweight
(linear projection only), treating it as ballast for distribution shift.

\paragraph{Reconstruction-error methods.}
DIRE~\cite{wang2023dire} and AEROBLADE~\cite{ricker2024aeroblade} detect
diffusion images via reconstruction error under a diffusion model or its
autoencoder. These require running (parts of) a generator at inference; we
exclude them from GRAFT's core to preserve local, CPU-class deployment, and
treat them as an optional plug-in branch in the released system.

\paragraph{Benchmarks and robustness limits.}
GenImage~\cite{zhu2023genimage} standardizes the cross-generator protocol we
adopt; Chameleon~\cite{yan2025sanity} stresses in-the-wild difficulty and its
companion method AIDE demonstrates mixture-of-experts hybrids; Community
Forensics~\cite{park2025community} scales generator diversity in training.
Saberi et al.~\cite{saberi2024robustness} give fundamental limits and
practical attacks; we therefore report perturbed metrics as first-class
results and position GRAFT's claims within, not against, those limits.

\section{Method}
\label{sec:method}
Given image $x \in [0,1]^{3\times H\times W}$, GRAFT outputs
$\hat p = P(y{=}\fake \mid x)$, per-branch probabilities, and gate weights.
Branches consume \emph{native-resolution} crops $c \in
[0,1]^{3\times S\times S}$ (default $S{=}256$); only the semantic branch sees
a resized global view. Nothing the forensic branches consume is ever
resampled (upscaling occurs only if $\min(H,W)<S$).

\subsection{Gradient-field branch}
\label{sec:gradient}
The branch input is a $17$-channel stack $\Phi(c)$ of gradient-domain maps.

\paragraph{(i) SRM residuals.} Four fixed $5{\times}5$ high-pass kernels
$\{w_k\}_{k=1}^{4}$ (the KV kernel, a Laplacian, and Scharr-normalized
horizontal/vertical derivative kernels), applied depthwise:
$r^{(k)}_{ch} = w_k * c_{ch}$, giving $12$ channels that suppress content and
retain synthesis noise~\cite{fridrich2012rich}.

\paragraph{(ii) Up-sampling (NPR-style) residual.} With bilinear operators
$D_{\downarrow2}, U_{\uparrow2}$,
\begin{equation}
r_{\mathrm{npr}} = c - U_{\uparrow2}\!\left(D_{\downarrow2}(c)\right),
\end{equation}
isolating traces of the up-sampling stages ubiquitous in generative
decoders~\cite{tan2024npr} ($3$ channels).

\paragraph{(iii) Gradient magnitude and coherence.} Let $I$ be the luma of
$c$ and $g = (g_x, g_y) = (\mathrm{Sch}_x * I,\ \mathrm{Sch}_y * I)$ its
Scharr gradient field. With a $7{\times}7$ box filter $B$ we form the
structure tensor $J = B * (g g^\top)$ and its \emph{coherence}
\begin{equation}
\kappa \;=\; \frac{\sqrt{(J_{11}-J_{22})^2 + 4J_{12}^2}}
{J_{11}+J_{22}+\varepsilon} \;\in\; [0,1],
\label{eq:coherence}
\end{equation}
the normalized eigenvalue anisotropy of local gradient orientation. Camera
imagery exhibits characteristic coherence statistics induced by optics,
demosaicing, and natural texture; synthesized micro-texture --- produced by
up-sampling stacks and attention-mediated decoders --- exhibits anomalous
local gradient alignment. To our knowledge, coherence of the gradient field
has not previously been used as an explicit input representation for
generated-image detection; we ablate its contribution in
\S\ref{sec:ablations}. Together with $|g|$ this adds $2$ channels.

A ResNet-18~\cite{he2016resnet} encoder (first convolution widened to $17$
input channels, ${\sim}11$M parameters) maps $\Phi(c)$ to
$f_{\mathrm{grad}} \in \mathbb{R}^{512}$. Optionally, an LGrad-style
channel~\cite{tan2023lgrad} --- the input-gradient of a frozen pretrained
classifier --- can be concatenated; it is disabled in the default
configuration to keep the fast tier download-free.

\subsection{Frequency branch}
For luma spectrum $F = \mathcal{F}(I)$ and annuli $A_\rho$ of normalized
radius $\rho$, the azimuthally averaged log-power profile is
\begin{equation}
P(\rho) = \tfrac{1}{|A_\rho|}\!\!\sum_{(u,v)\in A_\rho}\!\!
\log\!\left(1+|F(u,v)|\right),\quad \rho \in \{1,\dots,128\},
\end{equation}
standardized per image and concatenated with four band-energy statistics; a
two-layer MLP yields $f_{\mathrm{freq}} \in \mathbb{R}^{256}$
\cite{durall2020watch,frank2020leveraging}.

\subsection{Frozen semantic branch}
A CLIP ViT-B/16 image encoder~\cite{radford2021clip}, weights frozen, maps
the resized global view to features projected by a single trainable
linear layer to $f_{\mathrm{sem}} \in \mathbb{R}^{768}$. Freezing is a design
requirement, not a compute convenience: fine-tuning collapses the
cross-generator structure of the space~\cite{ojha2023universal}.

\subsection{Gated fusion, deep supervision, calibration}
\label{sec:fusion}
Each branch feature is projected to a shared space,
$z_b = \mathrm{GELU}(\mathrm{LN}(W_b f_b))$, and combined by a learned gate
$\alpha = \mathrm{softmax}(\gamma) \in \Delta^{B-1}$:
\begin{equation}
h = \textstyle\sum_b \alpha_b z_b, \qquad
s = \mathrm{MLP}(h), \qquad
\hat p = \sigma(s/T),
\end{equation}
with temperature $T$ fitted post hoc on validation
data~\cite{guo2017calibration}. Auxiliary heads give per-branch logits $s_b$;
the loss is
\begin{equation}
\mathcal{L} = \mathrm{BCE}(y, s) + \frac{\lambda}{B}\sum_b \mathrm{BCE}(y, s_b),
\qquad \lambda = 0.3 .
\end{equation}
Deep supervision stabilizes optimization and makes $(\alpha_b, \sigma(s_b))$
a per-image explanation of \emph{which evidence fired}. At inference, logits
are averaged over $K \le 8$ native crops (center plus corners), and $\hat p$
is mapped to \{\texttt{likely\_real}, \texttt{uncertain},
\texttt{likely\_ai}\} via thresholds $(\tau_r, \tau_f)$; abstention is an
explicit, reported outcome.

\subsection{Robustness-first training}
\label{sec:protocol}
Every training image passes a stochastic \emph{transport chain}: JPEG
re-encoding (quality $\sim\mathcal{U}\{30,\dots,95\}$, $p{=}0.5$), WebP
($p{=}0.1$), rescaling (factor $\mathcal{U}[0.5,1.5]$, $p{=}0.3$), Gaussian
blur ($p{=}0.15$), sensor-style noise ($p{=}0.1$), and horizontal flip ---
before random cropping. This follows the augmentation-for-generalization
insight of~\cite{wang2020cnn}, extended to the full re-encode--resize
pipeline that field images actually traverse, and paired with
\emph{mandatory} perturbed-set reporting at evaluation.

\section{Experimental Setup}
\paragraph{Datasets and protocol.} Primary: GenImage~\cite{zhu2023genimage}
under its strictest split --- train on Stable Diffusion v1.4 fakes plus
ImageNet reals \emph{only}; test on all eight subsets (ADM, BigGAN, GLIDE,
Midjourney, SD v1.4, SD v1.5, VQDM, Wukong). Secondary: the in-the-wild
Chameleon benchmark~\cite{yan2025sanity}, and a held-out set of
post-2025 commercial generators collected for the released
system \tofill{enumerate final generator list and collection dates}.

\paragraph{Baselines.} CNNDetection~\cite{wang2020cnn},
LGrad~\cite{tan2023lgrad}, NPR~\cite{tan2024npr},
UniversalFakeDetect~\cite{ojha2023universal}, DIRE~\cite{wang2023dire},
FatFormer~\cite{liu2024fatformer}, and AIDE~\cite{yan2025sanity}, all
retrained or re-evaluated under the identical protocol and identical
perturbations using released code where available
\tofill{note any baseline that could not be reproduced}.

\paragraph{Metrics.} AUROC and AP (ranking quality); accuracy at $0.5$;
TPR at $5\%$ FPR (the operationally relevant regime --- falsely accusing
real photographs is the costly error); expected calibration error (ECE);
and abstention rate at $(\tau_r,\tau_f) = (0.35, 0.65)$.

\paragraph{Implementation.} PyTorch; single NVIDIA A100 40GB; AdamW
(lr $2{\times}10^{-4}$, weight decay $0.05$), cosine schedule with $500$-step
warmup, batch $64$, $10$ epochs, mixed precision; crop $S{=}256$; the
semantic backbone contributes no gradients. The fast tier (gradient +
frequency) has ${\sim}12$M trainable parameters; the accurate tier adds the
frozen ViT-B ($+86$M frozen, $+0.6$M trainable). Full configuration files
ship with the code.

\section{Results}
\label{sec:results}
\emph{All numbers below are pending; the tables define the exact deliverable
of the experimental phase and will be populated only by runs reproducible
from the released harness.}

\begin{table}[t]
\centering\small
\caption{Cross-generator generalization (GenImage; train: SD v1.4 only).
AUROC per test subset. \tofill{complete after training runs}}
\begin{tabular}{l c c c}
\hline
Method & Mean & Worst subset & TPR@5\%FPR \\
\hline
CNNDetection~\cite{wang2020cnn} & \tofill{} & \tofill{} & \tofill{} \\
LGrad~\cite{tan2023lgrad}       & \tofill{} & \tofill{} & \tofill{} \\
NPR~\cite{tan2024npr}           & \tofill{} & \tofill{} & \tofill{} \\
UnivFD~\cite{ojha2023universal} & \tofill{} & \tofill{} & \tofill{} \\
DIRE~\cite{wang2023dire}        & \tofill{} & \tofill{} & \tofill{} \\
FatFormer~\cite{liu2024fatformer}& \tofill{} & \tofill{} & \tofill{} \\
AIDE~\cite{yan2025sanity}       & \tofill{} & \tofill{} & \tofill{} \\
\hline
GRAFT (fast)     & \tofill{} & \tofill{} & \tofill{} \\
GRAFT (accurate) & \tofill{} & \tofill{} & \tofill{} \\
\hline
\end{tabular}
\label{tab:main}
\end{table}

\begin{table}[t]
\centering\small
\caption{Transport robustness: mean AUROC over the eight GenImage subsets
under perturbation. \tofill{complete after training runs}}
\begin{tabular}{l c c c c}
\hline
Method & clean & JPEG50 & down$\times$0.5 & blur $\sigma{=}1$ \\
\hline
Best low-level baseline & \tofill{} & \tofill{} & \tofill{} & \tofill{} \\
Best CLIP baseline      & \tofill{} & \tofill{} & \tofill{} & \tofill{} \\
GRAFT (accurate)        & \tofill{} & \tofill{} & \tofill{} & \tofill{} \\
\hline
\end{tabular}
\label{tab:robust}
\end{table}

\subsection{Ablations}
\label{sec:ablations}
Planned and implemented in the harness: (a) remove each branch;
(b) remove the coherence channel $\kappa$ (Eq.~\ref{eq:coherence}) alone;
(c) replace native-resolution crops with global resize to $256^2$;
(d) disable transport-chain augmentation; (e) fine-tune vs.\ freeze the
semantic backbone; (f) gate analysis: distribution of $\alpha$ across clean
vs.\ perturbed inputs; (g) calibration on/off (ECE, abstention).
\tofill{ablation table}

\section{Limitations and Ethics}
GRAFT operates within known fundamental limits: sufficiently strong
adversarial post-processing can defeat any passive
detector~\cite{saberi2024robustness}, and the generator population is
non-stationary --- claims here are snapshots, which is why the release
couples the paper to a continuously refreshed evaluation harness rather than
a frozen artifact. Scores are probabilistic signals; we explicitly caution
against use as sole evidence in legal, journalistic, or academic-integrity
decisions, and the abstention band exists to make ``the model does not know''
a first-class output. Partially generated images (inpainting, AI-edited
photographs) and localization are out of scope for this work. Detection
research is dual-use: publishing which traces detectors read can guide
evasion; we follow the field's established norm that transparent, auditable
detection --- with honest robustness reporting --- serves defenders more than
secrecy serves attackers.

\section{Conclusion}
GRAFT unifies the two most generalization-robust families of forensic
evidence --- gradient-domain local statistics, extended with a
structure-tensor coherence representation, and frozen foundation-model
features --- under a gated, calibrated, robustness-first protocol, released
as a maintained open-source system. \tofill{one-sentence summary of empirical
findings once tables are complete.}

\paragraph{Reproducibility.} All code, configurations, manifests, and the
exact benchmark and perturbation harness are released at
\texttt{github.com/JitendraJha98/fakeradar}; weights are published with model
cards on the Hugging Face Hub.

\bibliographystyle{plain}
\bibliography{references}

\end{document}
